Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-10 paper/paper.tex
Remove the non-proof of thompsons construction for space

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to conducting formal
grammar theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname. By encoding well-established formal grammar concepts, we ensure
that our framework remains grounded in the
foundational principles of formal language theory while opening the door to
compositional formal verification of parsers.

@@

Working backwards, we can then show the traces of an NFA are weakly
equivalent to the traces of a DFA implementing a variant of Rabin and
Scott's classic powerset construction
\cite{rabinFiniteAutomataTheir1959}. Here we note that this is
\emph{only} a weak equivalence and not a strong equivalence, as the
DFA is unambiguous even if the NFA is not.
\begin{construction}[Determinization, \Agda]
  \label{cons:determinization}
  Given an NFA $N$, we construct a DFA $D$ such that
  $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
\end{construction}
\newcommand{\X}{\texttt{X}}
\newcommand{\Y}{\texttt{Y}}
\newcommand{\x}{\texttt{x}}
\newcommand{\y}{\texttt{y}}
\begin{proof}
  Define the states of $\D$ to be $\mathbb{P}_{\epsilon}(\N.\states)$ --- the type
  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
  from $\N$. Construct the initial state of $\D$ as the
  $\epsilon$-closure of $\N.\init$. Lastly, define the transition function
  of $\D$ to send the subset $X$ under the character $\c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labeled with the character $\c$.

@@

Finally, given any regular expression we can construct a
\emph{strongly} equivalent NFA. While only weak equivalence is
required to construct a parser, proving the strong equivalence shows
that other aspects of formal grammar theory are also verifiable in
\theoryabbv.
\newcommand{\R}{\texttt{R}}
\begin{construction}[Thompson's Construction, \Agda]
  \label{cons:thompson}
  Given a regular expression $\R$, we build an NFA $\N$ such that $\R$ is
  strongly equivalent to $\Trace_{N}(\N.\init)$.
\end{construction}
\begin{proof}
We use a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
\end{proof}
%% In this paper we will walk through
%% some of the cases of this proof. The remaining cases
%% follow
%% similar reasoning and the details may be found in the formalized artifact.

>>>>

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to formal
grammar theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname. By encoding well-established formal grammar concepts, we ensure
that our framework remains grounded in the
foundational principles of formal language theory while opening the door to
compositional formal verification of parsers.

@@

Working backwards, we can then show the traces of an NFA are weakly
equivalent to the traces of a DFA implementing a variant of Rabin and
Scott's classic powerset construction
\cite{rabinFiniteAutomataTheir1959}. Here we note that this is
\emph{only} a weak equivalence and not a strong equivalence, as the
DFA is unambiguous even if the NFA is not.
\begin{construction}[Determinization, \Agda]
  \label{cons:determinization}
  Given an NFA $N$, we construct a DFA $D$ such that
  $Parse_{N}$ is weakly equivalent to $Parse_{D}$.
\end{construction}
\newcommand{\X}{\texttt{X}}
\newcommand{\Y}{\texttt{Y}}
\newcommand{\x}{\texttt{x}}
\newcommand{\y}{\texttt{y}}
\begin{proof}
  Define the states of $\D$ to be $\mathbb{P}_{\epsilon}(\N.\states)$ --- the type
  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
  from $\N$. The initial state of $\D$ is the
  $\epsilon$-closure of $\N.\init$. Lastly, the transition function
  of $\D$ sends the subset $X$ under the character $\c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labeled with the character $\c$.

@@

Finally, given any regular expression we can construct a
\emph{strongly} equivalent NFA. While only weak equivalence is
required to construct a parser, proving the strong equivalence shows
that other aspects of formal grammar theory are also verifiable in
\theoryabbv.
\newcommand{\R}{\texttt{R}}
\begin{construction}[Thompson's Construction \cite{thompsonProgrammingTechniquesRegular1968}, \Agda]
  \label{cons:thompson}
  Given a regular expression $\R$, we build an NFA $\N$ such that $\R$ is
  strongly equivalent to $\Trace_{N}(\N.\init)$.
\end{construction}
% \begin{proof}
% We use a variant of Thompson's construction
% \cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
% \end{proof}
%% In this paper we will walk through
%% some of the cases of this proof. The remaining cases
%% follow
%% similar reasoning and the details may be found in the formalized artifact.
